\textbf{Differences in COVID-19 vaccination in the province of Ontario across Health Regions and socio-economic strata}

### 2023-07-10 manuscript_v6/main.tex
updated files, sent to Overleaf

\date{}
\begin{document}
\maketitle
\ifdefined\Shaded\renewenvironment{Shaded}{\begin{tcolorbox}[boxrule=0pt, breakable, borderline west={3pt}{0pt}{shadecolor}, sharp corners, frame hidden, enhanced, interior hidden]}{\end{tcolorbox}}\fi

@@

Briefly, the survey was deployed using random domain intercept
technology, a methodology where internet users whose device meta-data
indicated their presence in the province of Ontario had a random chance
of being redirected to the survey after they had clicked on a registered
but commercially inactive web link, or typed in a web address for a site
that was dormant but that was temporarily managed by the company that
administered the survey (RIWI
Corp)\textsuperscript{\protect\hyperlink{ref-sargent2022}{44}}. Users
then decided whether to anonymously participate in the survey, and those
that participated were able to exit the survey at any time. After the
survey closed, regardless if it was complete or incomplete, access was
denied to any further users with the same internet protocol address
(IP), effectively allowing each user only one opportunity to participate
in the survey. Users who indicated they were under the age of 16 were
exited from the survey without creating a record. Finally, the personal
identifier information from each user that participated in the survey
was automatically scrubbed and replaced by a unique ID.

@@

We used a logistic regression model to examine the impact of the Health
Regions in vaccination rates while considering the socio-economic
factors and months covered by the survey
(Table~\ref{tbl-descriptive-stats}) and certain interactions (Race and
Health Region and Race and income), as previous studies have shown that
socio-economic factors and their interactions are significant predictors
of intent of vaccination and vaccination
status\textsuperscript{\protect\hyperlink{ref-nguyen2022}{45}--\protect\hyperlink{ref-cnat2022a}{47}}.
Because we identified differences in representativity between the survey
data and the estimates from the 2016 Census for the variables considered
in the analysis, and used used an iterative proportional fitting
procedure
(\emph{raking})\textsuperscript{\protect\hyperlink{ref-deming1940}{48}}
to correct the data using data from the Census and Health Region
population totals; and fitted the regression model to the uncorrected
and corrected data. Details regarding the correction can be found in the
Appendix. All analyses were conducted in R 4.2.2 using the packages
\texttt{survey}\textsuperscript{\protect\hyperlink{ref-lumley2011}{49}},\texttt{tidyverse}\textsuperscript{\protect\hyperlink{ref-wickham2019}{50}},
\texttt{quarto}\textsuperscript{\protect\hyperlink{ref-quarto}{51}},
\texttt{modelsummary}\textsuperscript{\protect\hyperlink{ref-modelsummary}{52}},
and
\texttt{gtsummary}\textsuperscript{\protect\hyperlink{ref-gtsummary}{53}}.

@@

The COVID-19 pandemic has presented many significant challenges to the
healthcare systems of the Canadian
provinces\textsuperscript{\protect\hyperlink{ref-alami2021}{54}}. In the
case of Ontario, the recent adoption of the Health Region model at the
onset of the pandemic signified an additional level of complexity
because although PHO (through the 34 Public Health Units of the
province) was the lead in vaccine distribution, in reality the vaccine
rollout required active collaboration between PHO and OH in multiple
instances\textsuperscript{\protect\hyperlink{ref-ashcroft2023}{41}}.

In this study, we hypothesized that there were differences in the uptake
of the first dose of the COVID-19 vaccine between individuals living in
the different Health Regions of Ontario during late 2021 and early 2022.
Using the Health Regions as the base of our analysis was advantageous,
as these new Health Regions match closely the geographical boundaries of
the Health Regions that have been historically used to group Public
Health Units in the
province\textsuperscript{\protect\hyperlink{ref-bolotin2018}{55},\protect\hyperlink{ref-ohesi}{56}},
thus providing information with relevance at both the public health and
healthcare system level. Therefore, this analysis could provide an
overall assessment of intra-provincial disparities that might need to be
addressed moving forward by decision-makers, in order to ensure that in
the future, and specially in the event of a future public health
emergency, OH and PHO will able to collaboratively work in an efficient
way and thus ensure that the population of the province benefits in the
long term of the newly implemented Health Region model.

Our results show that indeed, there were differences in the uptake of
the first dose of the vaccine across Ontario in certain
socio-demographic groups. Specifically, those who identified as
Arab/Middle Eastern, Black, Latin American, having mixed racial or
ethnic background, or that belonged to other groups not explicitly
included in the survey (Southeast Asian, Filipino, West Asian, and
minority groups not identified elsewhere) had vaccination odds that were
between a third and a half of that of individuals that identified as
White or Caucasian (Figure~\ref{fig-models}). These results are
consistent with previous studies that have shown lower vaccination rates
in individuals with the same socio-demographic
characteristics\textsuperscript{\protect\hyperlink{ref-guay2022}{19}--\protect\hyperlink{ref-hussain2022}{21},\protect\hyperlink{ref-carter2022}{57}}.

Lower vaccine uptake in the socio-demographic groups indicated above may
be influenced in part, by vaccine hesitancy and refusal, which have been
associated in underrepresented Canadian individuals with concerns on
vaccine safety, effectiveness, and experiences of racial discrimination
in health
settings\textsuperscript{\protect\hyperlink{ref-cnat2022a}{47},\protect\hyperlink{ref-basta2022}{58}--\protect\hyperlink{ref-cnat2023}{60}}.
However, it has been shown that structural barriers also play an
important role in vaccination uptake. In the case of underrepresented
individuals, such barriers include complex scheduling systems, language
barriers, lack of adequate public transportation, and lack of accessible
vaccination
sites\textsuperscript{\protect\hyperlink{ref-njoku2021}{61}}. In this
regard, it is interesting to note that vaccination venues were scarce in
low socio-economic areas that had the highest burden of COVID-19 in
Toronto and other regions of Ontario around the time covered by the
survey\textsuperscript{\protect\hyperlink{ref-bogoch2022}{7},\protect\hyperlink{ref-iveniuk2021}{62}},
and that pharmacies in the Peel region (an area identified as a
``hotspot'' with high numbers of essential workers and multigenerational
households) could not keep up with vaccine
demand\textsuperscript{\protect\hyperlink{ref-gill2022}{63}}. This
suggests that the observed differences are associated with disparities
in vaccine access that were present during the period covered by the
survey.

@@

This result likely reflects in part the fact that individuals in
underrepresented groups tend to perform occupations that have been
deemed as ``essential'' in the context of the
pandemic\textsuperscript{\protect\hyperlink{ref-hawkins2020}{64},\protect\hyperlink{ref-ct2021}{65}},
which include workers in the areas of grocery stores, gas stations,
warehouses, distribution, and manufacturing, all being occupations for
which an income within the significant brackets identified in the
analysis is to be expected. From one side, individuals in essential
occupations in the province experienced higher rates of morbidity and
mortality during the first year of the
pandemic\textsuperscript{\protect\hyperlink{ref-rao2021}{66}}, but later
on, they had priority for COVID-19
vaccination\textsuperscript{\protect\hyperlink{ref-mishra2021}{67}}.
Additionally, it is known that vaccination uptake in these individuals
was encouraged by vaccination staff in certain parts of the
province\textsuperscript{\protect\hyperlink{ref-gill2022}{63}}. These
facts, combined with evidence of increased trends in vaccination in this
group
elsewhere\textsuperscript{\protect\hyperlink{ref-nguyen2021b}{68}},
suggest that in Ontario, the type of occupation of individuals in
underrepresented groups (which might have also affected their decision
to get a vaccine based on their knowledge of increased risk), played an
important role in the higher the odds of vaccination observed in these
individuals.

@@

First, in this case it is useful to analyze the data considering the
LHINs in each Health Region, because most studies in the literature
focused on Ontario use the LHINs as the base of their analyses. The West
Health Region covers the area previously occupied by the Hamilton
Niagara Haldimand Brant, South West, and Waterloo Wellington LHINs,
whereas the East Health Region covers the area of the former Champlain
and Central East LHINs. Previous research has identified health
disparities in these (mostly rural) regions, such as unequal
distribution of primary care providers, increased mortality, and low
pharmacist
availability\textsuperscript{\protect\hyperlink{ref-shah2019}{69}--\protect\hyperlink{ref-timony2022}{71}}.

Furthermore, there is an ongoing challenge for the health system of the
province with regard to personalized healthcare for marginalized
individuals. For example, the West Health Region has only two Aboriginal
Health Access Centres (community-led primary healthcare organizations
focused on First Nations, Métis, and Inuit communities) to provide care
to an estimated 100,000 Indigenous individuals living in the
area\textsuperscript{\protect\hyperlink{ref-ontariohealth}{72}}. Lack of
access to personalized healthcare affects individuals that may mistrust
the traditional healthcare system due to systemic racism or oppression,
which is known to be the case for Indigenous and Black individuals in
Canada, as these rationales have been associated with observed lower
vaccination rates among these
groups\textsuperscript{\protect\hyperlink{ref-smylie2022}{73},\protect\hyperlink{ref-eissa2021}{74}}.
Taken together, this suggests that healthcare disparities specific to
these underrepresented groups in certain parts of the province impacted
vaccination uptake, and highlights the need of investments in the Health
Regions focused on resources, infrastructure, and specially personnel
that can deliver personalized care to marginalized communities, as it
has been shown that such efforts have improved trust in vaccination in
underrepresented groups
elsewhere\textsuperscript{\protect\hyperlink{ref-schafferderoo2020}{75}}.

@@

Additionally, our analysis did not consider the North West and North
East Health Regions, due to the low number of entries from these areas
in the survey (Figure~\ref{fig-map}). Low representation is expected as
these regions as they only account for 5\% of the total population of
Ontario. However, these areas have the highest proportion of Indigenous
inhabitants\textsuperscript{\protect\hyperlink{ref-ontariohealth}{72}}.
In the context of personalized care, there is a need to collect data
that focuses on these Health Regions where additional health disparities
might be present and possibly understudied.

The results in this study are based on self-reported data, where bias
might be present. However, in the context of COVID-19, it has been shown
that good agreement exists between self-reported and documented
vaccination
status\textsuperscript{\protect\hyperlink{ref-stephenson2022}{76}}, we
believe that our data was able to provide a valid sample of vaccination
uptake in the province. This is supported by the statistically
significant higher vaccination odds that were identified for January of
2022 in the model, which are consistent with province-wide trends
reported by Public Health Ontario (which show a 4\% increase between
early December and January, in contrast to a 2.5\% increase between
October and
November\textsuperscript{\protect\hyperlink{ref-ontario-covid}{77}});
however, the short time window constitutes essentially a ``snapshot''
view of the evolution of the disease, and additional data would be
needed to obtain estimates per racial/ethnic group over time across all
Health Regions that can help inform the existence of other health
disparities.

@@

\leavevmode\vadjust pre{\hypertarget{ref-nguyen2022}{}}%
\CSLLeftMargin{45. }%
\CSLRightInline{Nguyen KH, Anneser E, Toppo A, Allen JD, Parott JS,
Corlin L. Disparities in national and state estimates of {COVID}-19
vaccination receipt and intent to vaccinate by race/ethnicity, income,
and age group among adults~\(\geq\)~18~years, united states.
\emph{Vaccine}. 2022;40(1):107-113.
doi:\href{https://doi.org/10.1016/j.vaccine.2021.11.040}{10.1016/j.vaccine.2021.11.040}}

\leavevmode\vadjust pre{\hypertarget{ref-shih2021}{}}%
\CSLLeftMargin{46. }%
\CSLRightInline{Shih SF, Wagner AL, Masters NB, Prosser LA, Lu Y,
Zikmund-Fisher BJ. Vaccine hesitancy and rejection of a vaccine for the
novel coronavirus in the united states. \emph{Frontiers in Immunology}.
2021;12.
doi:\href{https://doi.org/10.3389/fimmu.2021.558270}{10.3389/fimmu.2021.558270}}

\leavevmode\vadjust pre{\hypertarget{ref-cnat2022a}{}}%
\CSLLeftMargin{47. }%
\CSLRightInline{Cénat JM, Noorishad PG, Farahi SMMM, et al. Prevalence
and factors related to {COVID}-19 vaccine hesitancy and unwillingness in
canada: A systematic review and meta-analysis. \emph{Journal of Medical
Virology}. 2022;95(1).
doi:\href{https://doi.org/10.1002/jmv.28156}{10.1002/jmv.28156}}

\leavevmode\vadjust pre{\hypertarget{ref-deming1940}{}}%
\CSLLeftMargin{48. }%
\CSLRightInline{Deming WE, Stephan FF. On a least squares adjustment of
a sampled frequency table when the expected marginal totals are known.
\emph{The Annals of Mathematical Statistics}. 1940;11(4):427-444.
doi:\href{https://doi.org/10.1214/aoms/1177731829}{10.1214/aoms/1177731829}}

\leavevmode\vadjust pre{\hypertarget{ref-lumley2011}{}}%
\CSLLeftMargin{49. }%
\CSLRightInline{Lumley T. \emph{Complex Surveys}. John Wiley \& Sons;
2011.}

\leavevmode\vadjust pre{\hypertarget{ref-wickham2019}{}}%
\CSLLeftMargin{50. }%
\CSLRightInline{Wickham H, Averick M, Bryan J, et al. Welcome to the
{tidyverse}. \emph{Journal of Open Source Software}. 2019;4(43):1686.
doi:\href{https://doi.org/10.21105/joss.01686}{10.21105/joss.01686}}

\leavevmode\vadjust pre{\hypertarget{ref-quarto}{}}%
\CSLLeftMargin{51. }%
\CSLRightInline{Allaire J. \emph{Quarto: R Interface to 'Quarto'
Markdown Publishing System}.; 2022.
\url{https://CRAN.R-project.org/package=quarto}}

\leavevmode\vadjust pre{\hypertarget{ref-modelsummary}{}}%
\CSLLeftMargin{52. }%
\CSLRightInline{Arel-Bundock V. {modelsummary}: Data and model summaries
in {R}. \emph{Journal of Statistical Software}. 2022;103(1):1-23.
doi:\href{https://doi.org/10.18637/jss.v103.i01}{10.18637/jss.v103.i01}}

\leavevmode\vadjust pre{\hypertarget{ref-gtsummary}{}}%
\CSLLeftMargin{53. }%
\CSLRightInline{Sjoberg DD, Whiting K, Curry M, Lavery JA, Larmarange J.
Reproducible summary tables with the gtsummary package. \emph{{The R
Journal}}. 2021;13:570-580.
doi:\href{https://doi.org/10.32614/RJ-2021-053}{10.32614/RJ-2021-053}}

\leavevmode\vadjust pre{\hypertarget{ref-alami2021}{}}%
\CSLLeftMargin{54. }%
\CSLRightInline{Alami H, Lehoux P, Fleet R, et al. How can health
systems better prepare for the next pandemic? Lessons learned from the
management of {COVID}-19 in quebec (canada). \emph{Frontiers in Public
Health}. 2021;9.
doi:\href{https://doi.org/10.3389/fpubh.2021.671833}{10.3389/fpubh.2021.671833}}

\leavevmode\vadjust pre{\hypertarget{ref-bolotin2018}{}}%
\CSLLeftMargin{55. }%
\CSLRightInline{Bolotin JJAG Shelly AND Feld. Population-based estimate
of hepatitis c virus prevalence in ontario, canada. \emph{PLOS ONE}.
2018;13(1):1-10.
doi:\href{https://doi.org/10.1371/journal.pone.0191184}{10.1371/journal.pone.0191184}}

\leavevmode\vadjust pre{\hypertarget{ref-ohesi}{}}%
\CSLLeftMargin{56. }%
\CSLRightInline{Ontario HIV Epidemiology and Surveillance Initiative.
{T}echnical {N}otes: {H}ealth regions - {O}{H}{E}{S}{I} --- ohesi.ca.
Accessed July 7, 2023.
\url{https://www.ohesi.ca/technical-notes/health-regions/}}

\leavevmode\vadjust pre{\hypertarget{ref-carter2022}{}}%
\CSLLeftMargin{57. }%
\CSLRightInline{Carter MA, Biro S, Maier A, Shingler C, Guan TH.
{COVID}-19 vaccine uptake in southeastern ontario, canada: Monitoring
and addressing health inequities. \emph{Journal of Public Health
Management and Practice}. 2022;28(6):615-623.
doi:\href{https://doi.org/10.1097/phh.0000000000001565}{10.1097/phh.0000000000001565}}

\leavevmode\vadjust pre{\hypertarget{ref-basta2022}{}}%
\CSLLeftMargin{58. }%
\CSLRightInline{Basta NE, Sohel N, Sulis G, et al. Factors associated
with willingness to receive a {COVID}-19 vaccine among 23, 819 adults
aged 50 years or older: An analysis of the canadian longitudinal study
on aging. \emph{American Journal of Epidemiology}. 2022;191(6):987-998.
doi:\href{https://doi.org/10.1093/aje/kwac029}{10.1093/aje/kwac029}}

\leavevmode\vadjust pre{\hypertarget{ref-cnat2022b}{}}%
\CSLLeftMargin{59. }%
\CSLRightInline{Cénat JM, Noorishad PG, Bakombo SM, et al. A systematic
review on vaccine hesitancy in black communities in canada: Critical
issues and research failures. \emph{Vaccines}. 2022;10(11):1937.
doi:\href{https://doi.org/10.3390/vaccines10111937}{10.3390/vaccines10111937}}

\leavevmode\vadjust pre{\hypertarget{ref-cnat2023}{}}%
\CSLLeftMargin{60. }%
\CSLRightInline{Cénat JM, Farahi SMMM, Bakombo SM, et al. Vaccine
mistrust among black individuals in canada: The major role of health
literacy, conspiracy theories, and racial discrimination in the
healthcare system. \emph{Journal of Medical Virology}. 2023;95(4).
doi:\href{https://doi.org/10.1002/jmv.28738}{10.1002/jmv.28738}}

>>>>

\date{}
\begin{document}
\maketitle
\ifdefined\Shaded\renewenvironment{Shaded}{\begin{tcolorbox}[sharp corners, interior hidden, boxrule=0pt, borderline west={3pt}{0pt}{shadecolor}, enhanced, breakable, frame hidden]}{\end{tcolorbox}}\fi

@@

Briefly, the survey was deployed using Random Domain Intercept
Technology (RDIT), a methodology for internet surveys developed by the
company commissioned to run the survey (RIWI Corp., Toronto, Canada),
and that has been used in the area of public health research to examine
trends in vaccination
rates\textsuperscript{\protect\hyperlink{ref-sargent2022}{44}}, ratings
of care quality\textsuperscript{\protect\hyperlink{ref-kim2021}{45}},
and perceptions of vaccine
efficacy\textsuperscript{\protect\hyperlink{ref-seeman2010}{46}}. In the
case of the survey used for this study, internet users whose device
meta-data indicated their presence in the province of Ontario had a
random chance of being redirected to the survey after they had clicked
on a registered but commercially inactive web link, or after they had
typed in a web address for a site that was dormant but that was
temporarily managed by RIWI. Users then decided whether to anonymously
participate in the survey, and those that participated were able to exit
the survey at any time. After the survey closed, regardless if it was
complete or incomplete, access was denied to any further users with the
same internet protocol address (IP), effectively allowing each user only
one opportunity to participate in the survey. Users who indicated they
were under the age of 16 were exited from the survey without creating a
record. Finally, the personal identifier information from each user that
participated in the survey was automatically scrubbed and replaced by a
unique ID.

@@

We used a logistic regression model to examine the impact of the Health
Regions in vaccination rates while considering the socio-economic
factors and months covered by the survey
(Table~\ref{tbl-descriptive-stats}) and certain interactions (Race and
Health Region and Race and income), as previous studies have shown that
socio-economic factors and their interactions are significant predictors
of intent of vaccination and vaccination
status\textsuperscript{\protect\hyperlink{ref-nguyen2022}{47}--\protect\hyperlink{ref-cnat2022a}{49}}.
Because we identified differences in representativity between the survey
data and the estimates from the 2016 Census for the variables considered
in the analysis, and used used an iterative proportional fitting
procedure
(\emph{raking})\textsuperscript{\protect\hyperlink{ref-deming1940}{50}}
to correct the data using data from the Census and Health Region
population totals; and fitted the regression model to the uncorrected
and corrected data. Details regarding the correction can be found in the
Appendix. All analyses were conducted in R 4.2.2 using the packages
\texttt{survey}\textsuperscript{\protect\hyperlink{ref-lumley2011}{51}},\texttt{tidyverse}\textsuperscript{\protect\hyperlink{ref-wickham2019}{52}},
\texttt{quarto}\textsuperscript{\protect\hyperlink{ref-quarto}{53}},
\texttt{modelsummary}\textsuperscript{\protect\hyperlink{ref-modelsummary}{54}},
and
\texttt{gtsummary}\textsuperscript{\protect\hyperlink{ref-gtsummary}{55}}.

@@

The COVID-19 pandemic has presented many significant challenges to the
healthcare systems of the Canadian
provinces\textsuperscript{\protect\hyperlink{ref-alami2021}{56}}. In the
case of Ontario, the recent adoption of the Health Region model at the
onset of the pandemic signified an additional level of complexity
because although PHO (through the 34 Public Health Units of the
province) was the lead in vaccine distribution, in reality the vaccine
rollout required active collaboration between PHO and OH in multiple
instances\textsuperscript{\protect\hyperlink{ref-ashcroft2023}{41}}.

In this study, we hypothesized that there were differences in the uptake
of the first dose of the COVID-19 vaccine between individuals living in
the different Health Regions of Ontario during late 2021 and early 2022.
Using the Health Regions as the base of our analysis was advantageous,
as these new Health Regions match closely the geographical boundaries of
the Health Regions that have been historically used to group Public
Health Units in the
province\textsuperscript{\protect\hyperlink{ref-bolotin2018}{57},\protect\hyperlink{ref-ohesi}{58}},
thus providing information with relevance at both the public health and
healthcare system level. Therefore, this analysis could provide an
overall assessment of intra-provincial disparities that might need to be
addressed moving forward by decision-makers, in order to ensure that in
the future, and specially in the event of a future public health
emergency, OH and PHO will able to collaboratively work in an efficient
way and thus ensure that the population of the province benefits in the
long term of the newly implemented Health Region model.

Our results show that indeed, there were differences in the uptake of
the first dose of the vaccine across Ontario in certain
socio-demographic groups. Specifically, those who identified as
Arab/Middle Eastern, Black, Latin American, having mixed racial or
ethnic background, or that belonged to other groups not explicitly
included in the survey (Southeast Asian, Filipino, West Asian, and
minority groups not identified elsewhere) had vaccination odds that were
between a third and a half of that of individuals that identified as
White or Caucasian (Figure~\ref{fig-models}). These results are
consistent with previous studies that have shown lower vaccination rates
in individuals with the same socio-demographic
characteristics\textsuperscript{\protect\hyperlink{ref-guay2022}{19}--\protect\hyperlink{ref-hussain2022}{21},\protect\hyperlink{ref-carter2022}{59}}.

Lower vaccine uptake in the socio-demographic groups indicated above may
be influenced in part, by vaccine hesitancy and refusal, which have been
associated in underrepresented Canadian individuals with concerns on
vaccine safety, effectiveness, and experiences of racial discrimination
in health
settings\textsuperscript{\protect\hyperlink{ref-cnat2022a}{49},\protect\hyperlink{ref-basta2022}{60}--\protect\hyperlink{ref-cnat2023}{62}}.
However, it has been shown that structural barriers also play an
important role in vaccination uptake. In the case of underrepresented
individuals, such barriers include complex scheduling systems, language
barriers, lack of adequate public transportation, and lack of accessible
vaccination
sites\textsuperscript{\protect\hyperlink{ref-njoku2021}{63}}. In this
regard, it is interesting to note that vaccination venues were scarce in
low socio-economic areas that had the highest burden of COVID-19 in
Toronto and other regions of Ontario around the time covered by the
survey\textsuperscript{\protect\hyperlink{ref-bogoch2022}{7},\protect\hyperlink{ref-iveniuk2021}{64}},
and that pharmacies in the Peel region (an area identified as a
``hotspot'' with high numbers of essential workers and multigenerational
households) could not keep up with vaccine
demand\textsuperscript{\protect\hyperlink{ref-gill2022}{65}}. This
suggests that the observed differences are associated with disparities
in vaccine access that were present during the period covered by the
survey.

@@

This result likely reflects in part the fact that individuals in
underrepresented groups tend to perform occupations that have been
deemed as ``essential'' in the context of the
pandemic\textsuperscript{\protect\hyperlink{ref-hawkins2020}{66},\protect\hyperlink{ref-ct2021}{67}},
which include workers in the areas of grocery stores, gas stations,
warehouses, distribution, and manufacturing, all being occupations for
which an income within the significant brackets identified in the
analysis is to be expected. From one side, individuals in essential
occupations in the province experienced higher rates of morbidity and
mortality during the first year of the
pandemic\textsuperscript{\protect\hyperlink{ref-rao2021}{68}}, but later
on, they had priority for COVID-19
vaccination\textsuperscript{\protect\hyperlink{ref-mishra2021}{69}}.
Additionally, it is known that vaccination uptake in these individuals
was encouraged by vaccination staff in certain parts of the
province\textsuperscript{\protect\hyperlink{ref-gill2022}{65}}. These
facts, combined with evidence of increased trends in vaccination in this
group
elsewhere\textsuperscript{\protect\hyperlink{ref-nguyen2021b}{70}},
suggest that in Ontario, the type of occupation of individuals in
underrepresented groups (which might have also affected their decision
to get a vaccine based on their knowledge of increased risk), played an
important role in the higher the odds of vaccination observed in these
individuals.

@@

First, in this case it is useful to analyze the data considering the
LHINs in each Health Region, because most studies in the literature
focused on Ontario use the LHINs as the base of their analyses. The West
Health Region covers the area previously occupied by the Hamilton
Niagara Haldimand Brant, South West, and Waterloo Wellington LHINs,
whereas the East Health Region covers the area of the former Champlain
and Central East LHINs. Previous research has identified health
disparities in these (mostly rural) regions, such as unequal
distribution of primary care providers, increased mortality, and low
pharmacist
availability\textsuperscript{\protect\hyperlink{ref-shah2019}{71}--\protect\hyperlink{ref-timony2022}{73}}.

Furthermore, there is an ongoing challenge for the health system of the
province with regard to personalized healthcare for marginalized
individuals. For example, the West Health Region has only two Aboriginal
Health Access Centres (community-led primary healthcare organizations
focused on First Nations, Métis, and Inuit communities) to provide care
to an estimated 100,000 Indigenous individuals living in the
area\textsuperscript{\protect\hyperlink{ref-ontariohealth}{74}}. Lack of
access to personalized healthcare affects individuals that may mistrust
the traditional healthcare system due to systemic racism or oppression,
which is known to be the case for Indigenous and Black individuals in
Canada, as these rationales have been associated with observed lower
vaccination rates among these
groups\textsuperscript{\protect\hyperlink{ref-smylie2022}{75},\protect\hyperlink{ref-eissa2021}{76}}.
Taken together, this suggests that healthcare disparities specific to
these underrepresented groups in certain parts of the province impacted
vaccination uptake, and highlights the need of investments in the Health
Regions focused on resources, infrastructure, and specially personnel
that can deliver personalized care to marginalized communities, as it
has been shown that such efforts have improved trust in vaccination in
underrepresented groups
elsewhere\textsuperscript{\protect\hyperlink{ref-schafferderoo2020}{77}}.

@@

Additionally, our analysis did not consider the North West and North
East Health Regions, due to the low number of entries from these areas
in the survey (Figure~\ref{fig-map}). Low representation is expected as
these regions as they only account for 5\% of the total population of
Ontario. However, these areas have the highest proportion of Indigenous
inhabitants\textsuperscript{\protect\hyperlink{ref-ontariohealth}{74}}.
In the context of personalized care, there is a need to collect data
that focuses on these Health Regions where additional health disparities
might be present and possibly understudied.

The results in this study are based on self-reported data, where bias
might be present. However, in the context of COVID-19, it has been shown
that good agreement exists between self-reported and documented
vaccination
status\textsuperscript{\protect\hyperlink{ref-stephenson2022}{78}}, we
believe that our data was able to provide a valid sample of vaccination
uptake in the province. This is supported by the statistically
significant higher vaccination odds that were identified for January of
2022 in the model, which are consistent with province-wide trends
reported by Public Health Ontario (which show a 4\% increase between
early December and January, in contrast to a 2.5\% increase between
October and
November\textsuperscript{\protect\hyperlink{ref-ontario-covid}{79}});
however, the short time window constitutes essentially a ``snapshot''
view of the evolution of the disease, and additional data would be
needed to obtain estimates per racial/ethnic group over time across all
Health Regions that can help inform the existence of other health
disparities.

@@

\leavevmode\vadjust pre{\hypertarget{ref-kim2021}{}}%
\CSLLeftMargin{45. }%
\CSLRightInline{Kim JH, Bell GA, Ratcliffe HL, et al. Predictors of
patient-reported quality of care in low- and middle-income countries: A
four-country survey of person-centered care. \emph{International Journal
for Quality in Health Care}. 2021;33(3).
doi:\href{https://doi.org/10.1093/intqhc/mzab110}{10.1093/intqhc/mzab110}}

\leavevmode\vadjust pre{\hypertarget{ref-seeman2010}{}}%
\CSLLeftMargin{46. }%
\CSLRightInline{Seeman N, Ing A, Rizo C. Assessing and responding in
real time to online anti-vaccine sentiment during a flu pandemic.
\emph{Healthcare Quarterly}. 2010;13(sp):8-15.
doi:\href{https://doi.org/10.12927/hcq.2010.21923}{10.12927/hcq.2010.21923}}

\leavevmode\vadjust pre{\hypertarget{ref-nguyen2022}{}}%
\CSLLeftMargin{47. }%
\CSLRightInline{Nguyen KH, Anneser E, Toppo A, Allen JD, Parott JS,
Corlin L. Disparities in national and state estimates of {COVID}-19
vaccination receipt and intent to vaccinate by race/ethnicity, income,
and age group among adults~\(\geq\)~18~years, united states.
\emph{Vaccine}. 2022;40(1):107-113.
doi:\href{https://doi.org/10.1016/j.vaccine.2021.11.040}{10.1016/j.vaccine.2021.11.040}}

\leavevmode\vadjust pre{\hypertarget{ref-shih2021}{}}%
\CSLLeftMargin{48. }%
\CSLRightInline{Shih SF, Wagner AL, Masters NB, Prosser LA, Lu Y,
Zikmund-Fisher BJ. Vaccine hesitancy and rejection of a vaccine for the
novel coronavirus in the united states. \emph{Frontiers in Immunology}.
2021;12.
doi:\href{https://doi.org/10.3389/fimmu.2021.558270}{10.3389/fimmu.2021.558270}}

\leavevmode\vadjust pre{\hypertarget{ref-cnat2022a}{}}%
\CSLLeftMargin{49. }%
\CSLRightInline{Cénat JM, Noorishad PG, Farahi SMMM, et al. Prevalence
and factors related to {COVID}-19 vaccine hesitancy and unwillingness in
canada: A systematic review and meta-analysis. \emph{Journal of Medical
Virology}. 2022;95(1).
doi:\href{https://doi.org/10.1002/jmv.28156}{10.1002/jmv.28156}}

\leavevmode\vadjust pre{\hypertarget{ref-deming1940}{}}%
\CSLLeftMargin{50. }%
\CSLRightInline{Deming WE, Stephan FF. On a least squares adjustment of
a sampled frequency table when the expected marginal totals are known.
\emph{The Annals of Mathematical Statistics}. 1940;11(4):427-444.
doi:\href{https://doi.org/10.1214/aoms/1177731829}{10.1214/aoms/1177731829}}

\leavevmode\vadjust pre{\hypertarget{ref-lumley2011}{}}%
\CSLLeftMargin{51. }%
\CSLRightInline{Lumley T. \emph{Complex Surveys}. John Wiley \& Sons;
2011.}

\leavevmode\vadjust pre{\hypertarget{ref-wickham2019}{}}%
\CSLLeftMargin{52. }%
\CSLRightInline{Wickham H, Averick M, Bryan J, et al. Welcome to the
{tidyverse}. \emph{Journal of Open Source Software}. 2019;4(43):1686.
doi:\href{https://doi.org/10.21105/joss.01686}{10.21105/joss.01686}}

\leavevmode\vadjust pre{\hypertarget{ref-quarto}{}}%
\CSLLeftMargin{53. }%
\CSLRightInline{Allaire J. \emph{Quarto: R Interface to 'Quarto'
Markdown Publishing System}.; 2022.
\url{https://CRAN.R-project.org/package=quarto}}

\leavevmode\vadjust pre{\hypertarget{ref-modelsummary}{}}%
\CSLLeftMargin{54. }%
\CSLRightInline{Arel-Bundock V. {modelsummary}: Data and model summaries
in {R}. \emph{Journal of Statistical Software}. 2022;103(1):1-23.
doi:\href{https://doi.org/10.18637/jss.v103.i01}{10.18637/jss.v103.i01}}

\leavevmode\vadjust pre{\hypertarget{ref-gtsummary}{}}%
\CSLLeftMargin{55. }%
\CSLRightInline{Sjoberg DD, Whiting K, Curry M, Lavery JA, Larmarange J.
Reproducible summary tables with the gtsummary package. \emph{{The R
Journal}}. 2021;13:570-580.
doi:\href{https://doi.org/10.32614/RJ-2021-053}{10.32614/RJ-2021-053}}

\leavevmode\vadjust pre{\hypertarget{ref-alami2021}{}}%
\CSLLeftMargin{56. }%
\CSLRightInline{Alami H, Lehoux P, Fleet R, et al. How can health
systems better prepare for the next pandemic? Lessons learned from the
management of {COVID}-19 in quebec (canada). \emph{Frontiers in Public
Health}. 2021;9.
doi:\href{https://doi.org/10.3389/fpubh.2021.671833}{10.3389/fpubh.2021.671833}}

\leavevmode\vadjust pre{\hypertarget{ref-bolotin2018}{}}%
\CSLLeftMargin{57. }%
\CSLRightInline{Bolotin JJAG Shelly AND Feld. Population-based estimate
of hepatitis c virus prevalence in ontario, canada. \emph{PLOS ONE}.
2018;13(1):1-10.
doi:\href{https://doi.org/10.1371/journal.pone.0191184}{10.1371/journal.pone.0191184}}

\leavevmode\vadjust pre{\hypertarget{ref-ohesi}{}}%
\CSLLeftMargin{58. }%
\CSLRightInline{Ontario HIV Epidemiology and Surveillance Initiative.
{T}echnical {N}otes: {H}ealth regions - {O}{H}{E}{S}{I} --- ohesi.ca.
Accessed July 7, 2023.
\url{https://www.ohesi.ca/technical-notes/health-regions/}}

\leavevmode\vadjust pre{\hypertarget{ref-carter2022}{}}%
\CSLLeftMargin{59. }%
\CSLRightInline{Carter MA, Biro S, Maier A, Shingler C, Guan TH.
{COVID}-19 vaccine uptake in southeastern ontario, canada: Monitoring
and addressing health inequities. \emph{Journal of Public Health
Management and Practice}. 2022;28(6):615-623.
doi:\href{https://doi.org/10.1097/phh.0000000000001565}{10.1097/phh.0000000000001565}}

\leavevmode\vadjust pre{\hypertarget{ref-basta2022}{}}%
\CSLLeftMargin{60. }%
\CSLRightInline{Basta NE, Sohel N, Sulis G, et al. Factors associated
with willingness to receive a {COVID}-19 vaccine among 23, 819 adults
aged 50 years or older: An analysis of the canadian longitudinal study
on aging. \emph{American Journal of Epidemiology}. 2022;191(6):987-998.
doi:\href{https://doi.org/10.1093/aje/kwac029}{10.1093/aje/kwac029}}

\leavevmode\vadjust pre{\hypertarget{ref-cnat2022b}{}}%
\CSLLeftMargin{61. }%
\CSLRightInline{Cénat JM, Noorishad PG, Bakombo SM, et al. A systematic
review on vaccine hesitancy in black communities in canada: Critical
issues and research failures. \emph{Vaccines}. 2022;10(11):1937.
doi:\href{https://doi.org/10.3390/vaccines10111937}{10.3390/vaccines10111937}}
